\pkg{dataMaid}: your assistant for data cleaning in \proglang{R}

### 2017-01-24 latex/article_vol2.Rnw
Bibliography

  We describe an \proglang{R} package which implements an extensive
  and customizeable suite of quality assessment tools that can be applied to
  a dataset in order to identify potential problems in its
  variables. The results can be presented in an auto-generated,
  non-technical, stand-alone overview document, intended to be perused
  by an investigator with an understanding of the variables in the
  data, but not necessarily knowledge of \R{R}. Thereby, \pkg{dataMaid}
  aids the dialogue between data
  analysts and field experts, while also providing easy
  documentation of reproducible data cleaning steps and data quality
  control. Moreover, the \pkg{dataMaid} solution changes the data
  cleaning process from the usual ad hoc approach to a systematic,
  well-documented endeavor.  \pkg{dataMaid} also provides a suite of
  more typical \proglang{R} tools for interactive data quality
  assessment and -cleaning, where the data inspections all live and die
  within the \proglang{R} console.

@@

Online tools such as OpenRefine (\url{http://openrefine.org/}) and
\proglang{R}-packages such as \pkg{plyr}, and \pkg{data.table} have
made data wrangling a lot easier, but only a handful of packages such
as \pkg{editrules}, \pkg{validate}, \pkg{DataCombine}, and
\pkg{janitor} attempt to implement systematic, reproducible data
cleaning.  These packages use different approaches for data cleaning:
\pkg{editrules} and \pkg{validate} provide frameworks for setting up
and checking constraints on the variables, while \pkg{DataCombine} and
\pkg{janitor} both provide a few functions for identifying problems
(e.g, duplicates, dates coded as numbers, etc.) in data.

@@

This manuscript is structured as follows: First, in Section
\ref{sec:usingdataMaid}, we introduce the main representative of the first
paradigm, namely the \R{clean()} function, which generates data
cleaning overview documents. In the \pkg{dataMaid} package, we have
provided a number of default cleaning steps that cover the data
cleaning challenges, we find to be most common. Next, in Section
\ref{sec:interactiveCleanR}, we present the interactive mode of \pkg{dataMaid}, as motivated
by the second paradigm above. But, as any data analyst knows,
every dataset is different, and some datasets might include problems
that cannot be detected by our data checking functions. In Section
\ref{sec:extending}, we turn to the question of how \pkg{dataMaid} extensions
can be made, such that they are integrable with the \R{clean()}
function and with the other tools available in \pkg{dataMaid}.  At last,
in Section \ref{sec:specificExamples}, we discuss a number of examples of
specific data cleaning challenges and how \pkg{dataMaid} can be used to
solve them.

@@

The \R{clean()} function is the primary workhorse of \pkg{dataMaid} and
this is the only function needed if a standard battery
of tests are used to generate data cleaning summaries. The data
cleaning summary itself is an overview document, intended for reading
by humans, in either pdf or html format. Appendix \ref{sec:appendix1}
provides an example of a data cleaning output document, produced by
calling \R{clean()} on the dataset \R{toyData} available in
\pkg{dataMaid}. The first two pages of this data cleaning output are
shown in Figure~\ref{fig:example1}. \R{toyData} is a very
small ($15$ rows of $6$ variables), artificial dataset which was created to
illustrate the main capabilities of \pkg{dataMaid}. The following
commands load the dataset and produce the cleaning output:

\begin{figure}[tb]
\begin{center}
\frame{\includegraphics[width=7.5cm,page=2]{dataMaid_toyData.pdf}}
\frame{\includegraphics[width=7.5cm,page=3]{dataMaid_toyData.pdf}}
%\includepdf[pages={2}, pagecommand={}]{dataMaid_testData.pdf}
\end{center}
\label{fig:example1}
\caption{Example output from running \R{clean()} on the \R{toyData}
  dataset. First a summary of the full dataset is given and then
  type-dependent information on each variable is given in a table and
  a graph. Larger versions of the pages can be seen in
  Appendix~\ref{sec:appendix1}.}
\end{figure}

@@

By default, an \proglang{R} markdown file and a rendered pdf overview
document is produced, saved to the disc (in the working directory) and
opened for immediate inspection. Turning to Figure~\ref{fig:example1},
we see that such a data cleaning output document consists of two
parts. First, an overview of what was done is presented under the
title \textit{Data cleaning summary}. Secondly, each variable in the
dataset is presented in turn using (up to) three tools in the
\textit{Variable list}: A table summarizing key features of the
variable, a figure visualizing its distribution and a
list of flagged issues, if any. For instance, in the \R{numeric}-type variable
\R{var2} from \R{toyData}, \R{clean()} has identified two values that
are suspected to be miscoded missing values (\R{999} and \R{NaN}),
while two values were also flagged as potential outliers that should
be investigated more carefully.
%We can then return to Part
%1, the data cleaning summary, and inspect what sorts of checks were
%performed on this variable. \hl{hm, delete the last sentence? or use
%  different variable? Only these exact two checks were performed}.

Though the \R{clean()} function is very easy to use, it should not be
mistaken to be inflexible. A large number of function arguments allows
for the cleaning overview document to be molded according to the
user's needs. The most commonly used arguments are summarized in
Table~\ref{table.cleanFormals} and they are grouped according to the
part of the cleaning process they influence. In order to understand
this distinction, a glimpse of the inner structure of \R{clean()} is
shown in Figure~\ref{figure:cleanStructure}. Below, we present a few
examples on how to use the arguments from Table \ref{table.cleanFormals}
 to influence the output of a \R{clean()} call.

@@

\smallskip Control input variables and summary\\
\quad \R{useVar} & What variables should be used? & \R{NULL} (corresponding to all variables) \\
\quad \R{ordering} & Ordering of the variables in the data summary (as is or alphabetical) & \R{"asIs"} \\
\quad \R{onlyProblematic} & Should only variables flagged as problematic be included in the \textit{Variable list}? & \R{FALSE} \\
\quad \R{listChecks} & Should an overview of what checks were performed be listed in the \textit{Data cleaning summary}? &  \R{TRUE} \\
\quad \R{preChecks} & What check functions should be called to determine whether a variable is suitable for summarization, visualization and checking? & \R{c("isKey", "isEmpty")} \\

\smallskip Control summarize, visualize, and check steps \\
\quad \R{mode} & What steps should be performed for each variable (out
                 of the three possibilities \textit{summarize},
                 \textit{visualize}, \textit{check})? &
                                                        \R{c("summarize", "visualize", "check")} (corresponding to all three steps) \\
\quad \R{smartNum} & Should numerical values with only a few unique
                     levels be flagged and treated as a factor variable? & \R{TRUE} \\
\quad \R{maxProbVals} & Maximum number of problematic values to print, if any are found in data checks & \R{Inf} \\
\quad \R{maxDecimals} & Maximum number of decimals to print for numeric values in the variable list & \R{2} \\
\quad \R{twoCol} & Should the summary table and visualizations be placed side-by-side (in two columns)? & \R{TRUE} \\

\smallskip Control output and post-processing \\
\quad \R{output} & Type of output file to be produced (html, or pdf) & \R{"pdf"} \\
\quad \R{render} & Should the output file be rendered from markdown? & \R{TRUE} \\
\quad \R{openResult} & If a  pdf/html file is rendered, should it
                       automatically open afterwards, and if not,
                       should the \R{rmarkdown} file be opened? & \R{TRUE} \\
\quad \R{replace} & Overwrite an existing files with the same name? & \R{FALSE} \\

\hline
\end{tabular}
\caption{A selection of commonly used arguments to \R{clean()} separated into the parts they control.}
\label{table.cleanFormals}
\end{table}

% Below, we present a selection of these arguments in
% sections \hl{xx} and \hl{xxx}. We do this under two distinct headlines
% to emphasize that two levels of control are available when using
% \R{clean()}. Either, we can control the \hl{[something] - essentially,
%   this is everything but SVC-functions - } or we can control what
% quality assessments and descriptions each variable are subjected
% to.  \hl{segway to next bit or
%   just something to end the paragraph}

\begin{figure}[tb]
% \begin{description}
% \item[Input] A dataset
% 	\begin{itemize}
% 		\item This should be of type \R{data.frame} or \R{tibble} \hl{or data.table?}
% 	\end{itemize}
% \item[Create contents] For each variable in the dataset, do the following:
% 	\begin{description}
% 		\item[Stage 1:] Pre-checks
% 			\begin{itemize}
% 				\item Is the variable suitable for summarization, visualization and checks?
% 				\begin{description}
% 					\item[Yes:] Go to stage 2.
% 					\item[No:] Move on to the next variable.
% 				\end{description}
% 			\end{itemize}
% 		\item[Stage 2:] SVC-steps
% 			\begin{description}
% 				\item[Summarize] Call \R{summarize()} to produce a summary table describing the variable. What features enter this table depends on the data class of the variable.
% 				\item[Visualize]  Call \R{visualize()} to produce a plot visualizing the distribution of the variable.
% 				\item[Check]  Call \R{check()} to apply quality- and error checks to the variable. What checks are used depends on the data class of the variable.
% 			\end{description}
% 	\end{description}
% \item[Output] Files for a overview document are saved to the disc and possibly also opened.
% 	\begin{itemize}
% 		\item Always a \R{rmarkdown} (.Rmd) file
% 		\item Possible also a html or pdf file
% 	\end{itemize}
% \end{description}

@@

\caption{Schematic illustration of the stages undertaken when running
  \R{clean()}. Each variable is checked for eligibility before
  running \R{summarize()}, \R{visualize()}, and \R{check()}, and the
  resulting \proglang{R} markdown file may be rendered and opened.}
\label{figure:cleanStructure}
\end{figure}

@@

\subsection{Polishing off the arguments}
We begin with an example that is intended as an illustration of how \R{clean()} might
be used in the very first stages of data cleaning, where we are still uncertain
about exactly what is needed and how much times should be allocated to data
cleaning. At this stage, what is really needed, if a very rough idea of the extends of
errors in the dataset. In this scenario, we might wish to obtain a summary document in html format
that only contains the variables with potential problems, and with a limit of
maximum 10 printed potential errors for each variable we can write
(output not shown). Also, we can add the argument \R{replace=TRUE} in
order to force \R{clean()} to overwrite any existing files produced by
\R{clean()}, as we might run this function a few times in a row before deciding
exactly which cleaning overview we want. Using the \R{toyData} dataset as a guinea pig, we type:

@@

The final rendering of the generated markdown file is controlled by
the \R{render} and \R{openResult} arguments, which both default to
\R{TRUE}. \R{render} determines if the \proglang{R} markdown file
produced should be rendered using the \pkg{rmarkdown} package and
\R{openResult} decides whether the output html or pdf file should be
opened. The following command produces an \proglang{R} markdown file
containing the data cleaning summary information, but without
rendering nor opening the markdown file:

@@

\begin{table}
\centering
\begin{tabular}{p{0.35\linewidth} p{0.3\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth}
 p{0.01\linewidth} p{0.01\linewidth}}
  \hline
& Description &  \multicolumn{7}{c}{Variable classes} \\ \smallskip
 & &  C & F & I & L & B & N & D\\
  \hline \smallskip
  \textbf{\R{summaryFunction}s}  \smallskip \\
  \quad \R{centralValue} & Compute median or mode &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\
  \quad \R{countMissing} & Compute ratio of missing observations &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\
  \quad \R{minMax} & Find minimum and maximum values &   &  & $\times$ & &  & $\times$ & $\times$  \\
  \quad \R{quartiles} & Compute 1st and 3rd quartiles &    &  & $\times$ & &  & $\times$ & $\times$ \\
  \quad \R{uniqueValue} & Count number of unique values &   $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\
  \quad \R{variableType} & Data class of variable & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\
  \smallskip \\
 \textbf{\R{visualFunction}s} \smallskip \\
  \quad \R{basicVisual} & Histograms and barplots using base \proglang{R} graphics &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\
  \quad \R{standardVisual} & Histograms and barplots using ggplot2 &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\
  \smallskip \\
 \textbf{\R{checkFunction}s} \smallskip \\
 \quad \R{identifyCaseIssues} & Identify case issues &  $\times$ & $\times$ & & & & &  \\
 \quad \R{identifyLoners} & Identify levels with $<$ 6 obs. & $\times$ & $\times$ & & & & &  \\
 \quad \R{identifyMissing} & Identify miscoded missing values &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ &  \\
 \quad \R{identifyNums} & Identify misclassified numeric or integer variables & $\times$ & $\times$ & & & & &  \\
 \quad \R{identifyOutliers} & Identify outliers &  & & $\times$ & & $\times$ & $\times$ \\
 \quad \R{identifyOutliersTBStyle} & Identify outliers (Turkish Boxplot style) &  & & $\times$ & & $\times$ & $\times$ \\
 \quad \R{identifyWhitespace} & Identify prefixed and suffixed whitespace &  $\times$ & $\times$ & & $\times$ & & &  \\
 \quad \R{isCPR} & Identify Danish CPR numbers & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ &$\times$   \\
 \quad \R{isEmpty} & Check if the variable contains only a single value & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\
 \quad \R{isKey} & Check if the variable is a key & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$\smallskip   \\
 \hline
\end{tabular}
\caption{List of all summary functions (used in the summary table for
  each variable in the output), visual functions (used for  visualization of each variable), and
  check functions (used for data checks for each variable) currently implemented in \pkg{dataMaid}. The variable
  classes C, F, I, L, B, N, and D, refer to \R{character}, \R{factor},
  \R{integer}, \R{labelled}, \R{logical} (boolean), \R{numeric}, and \R{Date}, respectively.}
\label{table.SVCfunctions}
\end{table}

We will now move on to discuss how not only the structure of the
cleaning summary is manipulated, but also its very contents. This is
done through the summarize-visualize-check (SVC) step, as illustrated
in Figure \ref{figure:cleanStructure}.  \pkg{dataMaid} uses three
different types of functions for performing these steps, namely
\code{summaryFunction}s, \code{visualFunction}s and
\code{checkFunction}s.  By default, \R{clean()} runs selected summary,
visualization and check functions on each variable in the dataset, and
the exact choice of these functions depends on the classes of the
variables. For instance, though detection of outlier values might be
interesting for numerical variables, it holds little meaning for
factor variables, and therefore, numerical and factor variables need
different checks. Table~\ref{table.SVCfunctions} lists all available
summarize/visualize/check functions, but we can also use the
\R{allSummaryFunctions()}, \R{allVisualFunctions()}, and
\R{allCheckFunctions()} functions in \pkg{dataMaid} to print overview
lists in \proglang{R}. For example, the implemented
\R{summaryFunction}s are:

@@

Thus we can see, for example, that for \R{numeric}, \R{integer}, or \R{Date}
variables, \pkg{dataMaid} provides functions for adding summary information about
the minimum and maximum values, while all seven variable
classes dealt with in \pkg{dataMaid} have functions for central
tendency summaries (i.e., mode or median).

We can control what summaries and checks are applied for each variable type
through the \R{XXXSummaries} and \R{XXXChecks} arguments, where \R{XXX}
represents a variable type, e.g., \R{factorChecks} for factors,
\R{numericChecks} for numeric variables, etc. These arguments accept a
vector of summary- or check function names that should be run for a
particular variable type. The default values for each of these
arguments can be obtained through the \R{defaultXXXChecks()}
and \R{defaultXXXSummaries()} functions, where \R{XXX} again refers to
the variable type.

@@

We can change the summaries (and similarly the checks) by setting the
corresponding arguments when calling \R{clean()}. For example, to get
only the variable type and the central tendency listed in the summary
table of each variable we write

@@

In this particular case, where we specify the same set of summary
functions for each variable type, we can use the simpler argument
\R{allSummaries} which overrides the summary functions for all
variable types. Thus, the same result as above could be obtained with

@@

Similarly, the checks applied are set with the \R{XXXChecks}
arguments. The default checks being applied to a factor is

@@

Now, if we only wanted to apply the identify whitespace function for
factor variables, then we would need to set the \R{factorChecks} accordingly

@@

or we could remove checks for factors altogether by setting the
corresponding argument to \R{NULL}, in which case factor variables will
not be checked for any potential errors:

@@

As with \R{summaryFunction}s, a complete list of available
\R{checkFunction}s is obtained by calling
\R{allCheckFunctions()}. Note however, that \R{checkFunction}s have a
usage beyond the \R{XXXChecks} arguments, namely in the
\textit{pre-check} stage. In this stage, it is determined whether or
not each variable is suitable for the summarize/visualize/check (SVC)
steps. The functions used in the pre-check stage should be
\R{checkFunction}s that are applicable to all variable classes. The
default pre-checks, the functions \R{isKey()} and \R{isEmpty()}, check
whether a variable has unique values for all observations or only a
single value for all observations, respectively. If one of these
statements are true, the variable will not be subjected to the SVC
steps.  We can allow empty variables to move on to the SVC step by
only checking for keys in the pre-check step:

\begin{Schunk}
\begin{Sinput}
> clean(toyData, preChecks = "isKey")
\end{Sinput}
\end{Schunk}

Note that the data visualizations in the cleaning summary are also
controllable, though only a single function can be provided for all
variable types. If for instance, we wish to change the visualizations
from the default \pkg{ggplot2} style histograms and barplots to base
\proglang{R} histograms and barplots, we type
\begin{Schunk}
\begin{Sinput}
> clean(toyData, allVisuals = "basicVisual")
\end{Sinput}
\end{Schunk}

>>>>

  We describe an \proglang{R} package which implements an extensive
  and customizeable suite of quality assessment tools that can be applied to
  a dataset in order to identify potential problems in its
  variables. The results can be presented in an auto-generated,
  non-technical, stand-alone overview document, intended to be perused
  by an investigator with an understanding of the variables in the
  data, but not necessarily knowledge of \proglang{R}. Thereby, \pkg{dataMaid}
  aids the dialogue between data
  analysts and field experts, while also providing easy
  documentation of reproducible data cleaning steps and data quality
  control. Moreover, the \pkg{dataMaid} solution changes the data
  cleaning process from the usual ad hoc approach to a systematic,
  well-documented endeavor.  \pkg{dataMaid} also provides a suite of
  more typical \proglang{R} tools for interactive data quality
  assessment and -cleaning, where the data inspections all live and die
  within the \proglang{R} console.

@@

Online tools such as OpenRefine (\url{http://openrefine.org/}) and
\proglang{R}-packages such as \pkg{plyr} \citep{plyr}, and \pkg{data.table} have
made data wrangling a lot easier, but only a handful of packages such
as \pkg{editrules}, \pkg{validate}, \pkg{DataCombine}, and
\pkg{janitor} attempt to implement systematic, reproducible data
cleaning.  These packages use different approaches for data cleaning:
\pkg{editrules} and \pkg{validate} provide frameworks for setting up
and checking constraints on the variables, while \pkg{DataCombine} and
\pkg{janitor} both provide a few functions for identifying problems
(e.g, duplicates, dates coded as numbers, etc.) in data.

@@

This manuscript is structured as follows: First, in Section
\ref{sec:usingdataMaid}, we introduce the main representative of the first
paradigm, namely the \code{clean()} function, which generates data
cleaning overview documents. In the \pkg{dataMaid} package, we have
provided a number of default cleaning steps that cover the data
cleaning challenges, we find to be most common. Next, in Section
\ref{sec:interactiveCleanR}, we present the interactive mode of \pkg{dataMaid}, as motivated
by the second paradigm above. But, as any data analyst knows,
every dataset is different, and some datasets might include problems
that cannot be detected by our data checking functions. In Section
\ref{sec:extending}, we turn to the question of how \pkg{dataMaid} extensions
can be made, such that they are integrable with the \code{clean()}
function and with the other tools available in \pkg{dataMaid}.  At last,
in Section \ref{sec:specificExamples}, we discuss a number of examples of
specific data cleaning challenges and how \pkg{dataMaid} can be used to
solve them.

@@

The \code{clean()} function is the primary workhorse of \pkg{dataMaid} and
this is the only function needed if a standard battery
of tests are used to generate data cleaning summaries. The data
cleaning summary itself is an overview document, intended for reading
by humans, in either pdf or html format. Appendix \ref{sec:appendix1}
provides an example of a data cleaning output document, produced by
calling \code{clean()} on the dataset \code{toyData} available in
\pkg{dataMaid}. The first two pages of this data cleaning output are
shown in Figure~\ref{fig:example1}. \code{toyData} is a very
small ($15$ rows of $6$ variables), artificial dataset which was created to
illustrate the main capabilities of \pkg{dataMaid}. The following
commands load the dataset and produce the cleaning output:

\begin{figure}[tb]
\begin{center}
\frame{\includegraphics[width=7.5cm,page=2]{dataMaid_toyData.pdf}}
\frame{\includegraphics[width=7.5cm,page=3]{dataMaid_toyData.pdf}}
%\includepdf[pages={2}, pagecommand={}]{dataMaid_testData.pdf}
\end{center}
\label{fig:example1}
\caption{Example output from running \code{clean()} on the \code{toyData}
  dataset. First a summary of the full dataset is given and then
  type-dependent information on each variable is given in a table and
  a graph. Larger versions of the pages can be seen in
  Appendix~\ref{sec:appendix1}.}
\end{figure}

@@

By default, an \proglang{R} markdown file and a rendered pdf overview
document is produced, saved to the disc (in the working directory) and
opened for immediate inspection. Turning to Figure~\ref{fig:example1},
we see that such a data cleaning output document consists of two
parts. First, an overview of what was done is presented under the
title \textit{Data cleaning summary}. Secondly, each variable in the
dataset is presented in turn using (up to) three tools in the
\textit{Variable list}: A table summarizing key features of the
variable, a figure visualizing its distribution and a
list of flagged issues, if any. For instance, in the \code{numeric}-type variable
\code{var2} from \code{toyData}, \code{clean()} has identified two values that
are suspected to be miscoded missing values (\code{999} and \code{NaN}),
while two values were also flagged as potential outliers that should
be investigated more carefully.
%We can then return to Part
%1, the data cleaning summary, and inspect what sorts of checks were
%performed on this variable. \hl{hm, delete the last sentence? or use
%  different variable? Only these exact two checks were performed}.

Though the \code{clean()} function is very easy to use, it should not be
mistaken to be inflexible. A large number of function arguments allows
for the cleaning overview document to be molded according to the
user's needs. The most commonly used arguments are summarized in
Table~\ref{table.cleanFormals} and they are grouped according to the
part of the cleaning process they influence. In order to understand
this distinction, a glimpse of the inner structure of \code{clean()} is
shown in Figure~\ref{figure:cleanStructure}. Below, we present a few
examples on how to use the arguments from Table \ref{table.cleanFormals}
 to influence the output of a \code{clean()} call.

@@

\smallskip Control input variables and summary\\
\quad \code{useVar} & What variables should be used? & \code{NULL} (corresponding to all variables) \\
\quad \code{ordering} & Ordering of the variables in the data summary (as is or alphabetical) & \code{"asIs"} \\
\quad \code{onlyProblematic} & Should only variables flagged as problematic be included in the \textit{Variable list}? & \code{FALSE} \\
\quad \code{listChecks} & Should an overview of what checks were performed be listed in the \textit{Data cleaning summary}? &  \code{TRUE} \\
\quad \code{preChecks} & What check functions should be called to determine whether a variable is suitable for summarization, visualization and checking? & \code{c("isKey", "isEmpty")} \\

\smallskip Control summarize, visualize, and check steps \\
\quad \code{mode} & What steps should be performed for each variable (out
                 of the three possibilities \textit{summarize},
                 \textit{visualize}, \textit{check})? &
                                                        \code{c("summarize", "visualize", "check")} (corresponding to all three steps) \\
\quad \code{smartNum} & Should numerical values with only a few unique
                     levels be flagged and treated as a factor variable? & \code{TRUE} \\
\quad \code{maxProbVals} & Maximum number of problematic values to print, if any are found in data checks & \code{Inf} \\
\quad \code{maxDecimals} & Maximum number of decimals to print for numeric values in the variable list & \code{2} \\
\quad \code{twoCol} & Should the summary table and visualizations be placed side-by-side (in two columns)? & \code{TRUE} \\

\smallskip Control output and post-processing \\
\quad \code{output} & Type of output file to be produced (html, or pdf) & \code{"pdf"} \\
\quad \code{render} & Should the output file be rendered from markdown? & \code{TRUE} \\
\quad \code{openResult} & If a  pdf/html file is rendered, should it
                       automatically open afterwards, and if not,
                       should the \code{rmarkdown} file be opened? & \code{TRUE} \\
\quad \code{replace} & Overwrite an existing files with the same name? & \code{FALSE} \\

\hline
\end{tabular}
\caption{A selection of commonly used arguments to \code{clean()} separated into the parts they control.}
\label{table.cleanFormals}
\end{table}

% Below, we present a selection of these arguments in
% sections \hl{xx} and \hl{xxx}. We do this under two distinct headlines
% to emphasize that two levels of control are available when using
% \code{clean()}. Either, we can control the \hl{[something] - essentially,
%   this is everything but SVC-functions - } or we can control what
% quality assessments and descriptions each variable are subjected
% to.  \hl{segway to next bit or
%   just something to end the paragraph}

\begin{figure}[tb]
% \begin{description}
% \item[Input] A dataset
% 	\begin{itemize}
% 		\item This should be of type \code{data.frame} or \code{tibble} \hl{or data.table?}
% 	\end{itemize}
% \item[Create contents] For each variable in the dataset, do the following:
% 	\begin{description}
% 		\item[Stage 1:] Pre-checks
% 			\begin{itemize}
% 				\item Is the variable suitable for summarization, visualization and checks?
% 				\begin{description}
% 					\item[Yes:] Go to stage 2.
% 					\item[No:] Move on to the next variable.
% 				\end{description}
% 			\end{itemize}
% 		\item[Stage 2:] SVC-steps
% 			\begin{description}
% 				\item[Summarize] Call \code{summarize()} to produce a summary table describing the variable. What features enter this table depends on the data class of the variable.
% 				\item[Visualize]  Call \code{visualize()} to produce a plot visualizing the distribution of the variable.
% 				\item[Check]  Call \code{check()} to apply quality- and error checks to the variable. What checks are used depends on the data class of the variable.
% 			\end{description}
% 	\end{description}
% \item[Output] Files for a overview document are saved to the disc and possibly also opened.
% 	\begin{itemize}
% 		\item Always a \code{rmarkdown} (.Rmd) file
% 		\item Possible also a html or pdf file
% 	\end{itemize}
% \end{description}

@@

\caption{Schematic illustration of the stages undertaken when running
  \code{clean()}. Each variable is checked for eligibility before
  running \code{summarize()}, \code{visualize()}, and \code{check()}, and the
  resulting \proglang{R} markdown file may be rendered and opened.}
\label{figure:cleanStructure}
\end{figure}

@@

\subsection{Polishing off the arguments}
We begin with an example that is intended as an illustration of how \code{clean()} might
be used in the very first stages of data cleaning, where we are still uncertain
about exactly what is needed and how much times should be allocated to data
cleaning. At this stage, what is really needed, if a very rough idea of the extends of
errors in the dataset. In this scenario, we might wish to obtain a summary document in html format
that only contains the variables with potential problems, and with a limit of
maximum 10 printed potential errors for each variable we can write
(output not shown). Also, we can add the argument \code{replace=TRUE} in
order to force \code{clean()} to overwrite any existing files produced by
\code{clean()}, as we might run this function a few times in a row before deciding
exactly which cleaning overview we want. Using the \code{toyData} dataset as a guinea pig, we type:

@@

The final rendering of the generated markdown file is controlled by
the \code{render} and \code{openResult} arguments, which both default to
\code{TRUE}. \code{render} determines if the \proglang{R} markdown file
produced should be rendered using the \pkg{rmarkdown} package and
\code{openResult} decides whether the output html or pdf file should be
opened. The following command produces an \proglang{R} markdown file
containing the data cleaning summary information, but without
rendering nor opening the markdown file:

@@

\begin{table}
\centering
\begin{tabular}{p{0.35\linewidth} p{0.3\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth}
 p{0.01\linewidth} p{0.01\linewidth}}
  \hline
& Description &  \multicolumn{7}{c}{Variable classes} \\ \smallskip
 & &  C & F & I & L & B & N & D\\
  \hline \smallskip
  \textbf{\code{summaryFunction}s}  \smallskip \\
  \quad \code{centralValue} & Compute median or mode &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\
  \quad \code{countMissing} & Compute ratio of missing observations &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\
  \quad \code{minMax} & Find minimum and maximum values &   &  & $\times$ & &  & $\times$ & $\times$  \\
  \quad \code{quartiles} & Compute 1st and 3rd quartiles &    &  & $\times$ & &  & $\times$ & $\times$ \\
  \quad \code{uniqueValue} & Count number of unique values &   $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\
  \quad \code{variableType} & Data class of variable & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\
  \smallskip \\
 \textbf{\code{visualFunction}s} \smallskip \\
  \quad \code{basicVisual} & Histograms and barplots using base \proglang{R} graphics &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\
  \quad \code{standardVisual} & Histograms and barplots using ggplot2 &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\
  \smallskip \\
 \textbf{\code{checkFunction}s} \smallskip \\
 \quad \code{identifyCaseIssues} & Identify case issues &  $\times$ & $\times$ & & & & &  \\
 \quad \code{identifyLoners} & Identify levels with $<$ 6 obs. & $\times$ & $\times$ & & & & &  \\
 \quad \code{identifyMissing} & Identify miscoded missing values &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ &  \\
 \quad \code{identifyNums} & Identify misclassified numeric or integer variables & $\times$ & $\times$ & & & & &  \\
 \quad \code{identifyOutliers} & Identify outliers &  & & $\times$ & & $\times$ & $\times$ \\
 \quad \code{identifyOutliersTBStyle} & Identify outliers (Turkish Boxplot style) &  & & $\times$ & & $\times$ & $\times$ \\
 \quad \code{identifyWhitespace} & Identify prefixed and suffixed whitespace &  $\times$ & $\times$ & & $\times$ & & &  \\
 \quad \code{isCPR} & Identify Danish CPR numbers & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ &$\times$   \\
 \quad \code{isEmpty} & Check if the variable contains only a single value & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\
 \quad \code{isKey} & Check if the variable is a key & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$\smallskip   \\
 \hline
\end{tabular}
\caption{List of all summary functions (used in the summary table for
  each variable in the output), visual functions (used for  visualization of each variable), and
  check functions (used for data checks for each variable) currently implemented in \pkg{dataMaid}. The variable
  classes C, F, I, L, B, N, and D, refer to \code{character}, \code{factor},
  \code{integer}, \code{labelled}, \code{logical} (boolean), \code{numeric}, and \code{Date}, respectively.}
\label{table.SVCfunctions}
\end{table}

We will now move on to discuss how not only the structure of the
cleaning summary is manipulated, but also its very contents. This is
done through the summarize-visualize-check (SVC) step, as illustrated
in Figure \ref{figure:cleanStructure}.  \pkg{dataMaid} uses three
different types of functions for performing these steps, namely
\code{summaryFunction}s, \code{visualFunction}s and
\code{checkFunction}s.  By default, \code{clean()} runs selected summary,
visualization and check functions on each variable in the dataset, and
the exact choice of these functions depends on the classes of the
variables. For instance, though detection of outlier values might be
interesting for numerical variables, it holds little meaning for
factor variables, and therefore, numerical and factor variables need
different checks. Table~\ref{table.SVCfunctions} lists all available
summarize/visualize/check functions, but we can also use the
\code{allSummaryFunctions()}, \code{allVisualFunctions()}, and
\code{allCheckFunctions()} functions in \pkg{dataMaid} to print overview
lists in \proglang{R}. For example, the implemented
\code{summaryFunction}s are:

@@

Thus we can see, for example, that for \code{numeric}, \code{integer}, or \code{Date}
variables, \pkg{dataMaid} provides functions for adding summary information about
the minimum and maximum values, while all seven variable
classes dealt with in \pkg{dataMaid} have functions for central
tendency summaries (i.e., mode or median).

We can control what summaries and checks are applied for each variable type
through the \code{XXXSummaries} and \code{XXXChecks} arguments, where \code{XXX}
represents a variable type, e.g., \code{factorChecks} for factors,
\code{numericChecks} for numeric variables, etc. These arguments accept a
vector of summary- or check function names that should be run for a
particular variable type. The default values for each of these
arguments can be obtained through the \code{defaultXXXChecks()}
and \code{defaultXXXSummaries()} functions, where \code{XXX} again refers to
the variable type.

@@

We can change the summaries (and similarly the checks) by setting the
corresponding arguments when calling \code{clean()}. For example, to get
only the variable type and the central tendency listed in the summary
table of each variable we write

@@

In this particular case, where we specify the same set of summary
functions for each variable type, we can use the simpler argument
\code{allSummaries} which overrides the summary functions for all
variable types. Thus, the same result as above could be obtained with

@@

Similarly, the checks applied are set with the \code{XXXChecks}
arguments. The default checks being applied to a factor is

@@

Now, if we only wanted to apply the identify whitespace function for
factor variables, then we would need to set the \code{factorChecks} accordingly

@@

or we could remove checks for factors altogether by setting the
corresponding argument to \code{NULL}, in which case factor variables will
not be checked for any potential errors:

@@

As with \code{summaryFunction}s, a complete list of available
\code{checkFunction}s is obtained by calling
\code{allCheckFunctions()}. Note however, that \code{checkFunction}s have a
usage beyond the \code{XXXChecks} arguments, namely in the
\textit{pre-check} stage. In this stage, it is determined whether or
not each variable is suitable for the summarize/visualize/check (SVC)
steps. The functions used in the pre-check stage should be
\code{checkFunction}s that are applicable to all variable classes. The
default pre-checks, the functions \code{isKey()} and \code{isEmpty()}, check
whether a variable has unique values for all observations or only a
single value for all observations, respectively. If one of these
statements are true, the variable will not be subjected to the SVC
steps.  We can allow empty variables to move on to the SVC step by
only checking for keys in the pre-check step:

<<eval=FALSE>>=
clean(toyData, preChecks = "isKey")
@

Note that the data visualizations in the cleaning summary are also
controllable, though only a single function can be provided for all
variable types. If for instance, we wish to change the visualizations
from the default \pkg{ggplot2} style histograms and barplots to base
\proglang{R} histograms and barplots, we type

<<eval=FALSE>>=
clean(toyData, allVisuals = "basicVisual")
@
